---
title: "Evaluating the Impact of $100 Gift Cards on Vaccination Rates in Minnesota"
author: "Richard Bamattre"
output:
  html_document:
    df_print: paged
---

Looking at vaccine rates nationally and evaluating the impact of Minnesota Department of Health offering $100 gift card incentives for the first half of August 2021.

Studies like [this assessment of the vaccine lottery in Ohio](https://jamanetwork.com/journals/jama/fullarticle/2781792) are models for this approach.

[This study](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-021-01235-8) uses the fable package and gets more into the math.

**Note: This is an experiment to try out a method that I'm not very familiar with. It should not be construed as a reputable or peer-reviewed evaluation of vaccine incentives**

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE)

# load
library(tidycensus); census_api_key("0c5d9fddc67544f23d3fa524660c9d81d4be2c90")

# wrangle
library(tidyverse) # of course
library(janitor) # nice tables

# viz
library(ggthemes) # extra ggplot themes
library(gghighlight) # highlight key elements in ggplot

# model
library(tsibble) # time series tibbles/dataframes, required for fable
library(fable) # forecasting
library(feasts) # additional tools for forecasting
```

# Get data

## Data source

Originally used Johns Hopkins cleaned up data, but this only had cumulative doses. Switched to [Our World In Data](https://github.com/owid/covid-19-data/tree/master/public/data/vaccinations/), which downloads from the CDC for US Vaccinations.

This data is nice because vaccination rates per 100 or million are calculated.

```{r}
#vacc <- read_csv("https://github.com/govex/COVID-19/blob/master/data_tables/vaccine_data/us_data/time_series/people_vaccinated_us_timeline.csv?raw=true")

vacc <- read_csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/us_state_vaccinations.csv")

tabyl(vacc$location)
```

## Total vaccinations nationally

```{r}
vacc %>%
  filter(location == "United States") %>%
  ggplot(aes(x = date, y = daily_vaccinations_per_million)) +
    geom_line() +
    scale_y_continuous(labels = scales::comma) +
    labs(title = "Daily Vaccinations per Million, United States",
         y = "") +
    theme_fivethirtyeight()
```

### Looking at Minnesota's rate

```{r}
vacc %>%
  ggplot(aes(x = date, y = daily_vaccinations_per_million, group = location, 
             color = location)) +
    geom_line() +
    gghighlight(location %in% c("United States", "Minnesota", "Michigan",
                                "Wisconsin", "New York", "Alabama")) +
    labs(title = "Daily Vaccinations per Million Residents",
         y = "Vaccinations per million residents") +
    scale_y_continuous(labels = scales::comma) +
    theme_fivethirtyeight()

vacc %>%
  filter(location %in% c("United States", "Minnesota"),
         date > '2021-07-01') %>%
  ggplot(aes(x = date, y = daily_vaccinations_per_million, group = location, 
             color = location)) +
    geom_line() +
    theme_fivethirtyeight() +
    labs(title = "Daily Vaccinations per Million Residents",
         color = "")
```

Not sure why there is a dip on August 15 and another one a week later. Is this based on MN not reporting vaccinations on weekends? I would exclude the Sunday of August 15 just in case.

## Research Question

**Research Question: Was incentives introduced in August associated with a relative increase in vaccination rates in MN (compared to the US)?**

Incentives for shots done 7/30 - 8/15

Trying out an Interrupted Time Series (ITS) Approach

Using [this article](https://www.academicpedsjnl.net/article/S1876-2859(13)00210-6/fulltext) as a guide.

## Create time series dataframe

Create a dataframe to resemble that in Table 2 in the article. This has the coefficient necessary to do a ITS analysis.

```{r}
its <- vacc %>%
  filter(location %in% c("United States", "Minnesota"),
         date > '2021-07-14',
         date <= '2021-08-14') %>%
  select(location, date, daily_vaccinations_per_million) %>%
  pivot_wider(names_from = location, values_from = daily_vaccinations_per_million) %>%
  arrange(date) %>%
  mutate(diff = Minnesota - `United States`,
         prog = if_else(date >= '2021-07-30' & date <= '2021-08-15', 1, 0),
         time = row_number(),
         time_after = cumsum(prog))

its_ts <- its %>%
  as_tsibble(key = NULL, index = date)
```


```{r}
its %>%
  ggplot(aes(x = date)) +
    geom_point(aes(y = Minnesota), color = "red", shape = 1) +
    geom_smooth(aes(y = Minnesota, group = prog, color = factor(prog)), method = "lm",
                se = FALSE) +
    geom_point(aes(y = `United States`), color = "blue", shape = 1) +
    geom_smooth(aes(y = `United States`, group = prog, color = factor(prog)), method = "lm",
                se = FALSE) +
    labs(title = "Vaccination rates pre/post MN incentive",
         subtitle = "Minnesota (red) compared to US (blue)", y = "At least one vaccine per 100,000 people",
         color = "Program") +
    scale_y_continuous(labels = scales::comma) +
    theme_fivethirtyeight()
```


```{r}
its %>%
  mutate(Program = if_else(prog == 1, "Gift Cards", "No Gift Cards")) %>%
  ggplot(aes(x = date, y = diff)) +
    geom_point() +
    geom_smooth(aes(group = Program, color = factor(Program)), method = "lm",
                se = FALSE) +
    labs(title = "Vaccine rates: Minnesota minus the US",
         subtitle = "Before and during $100 gift cards", y = "Difference in vaccines per 100,000 people",
         color = "") +
    scale_y_continuous(labels = scales::comma) +
    theme_fivethirtyeight()
```


## Create model

Fit model using regression equation:

rate = 0 + time + program + time after program + e

```{r}
model <- its_ts %>%
  model(ARIMA(diff ~ time + prog + time_after))

report(model)
```

### Assess residuals

```{r}
model %>%
  gg_tsresiduals()
```

### Get full table of coefficients

```{r}
results <- model[[1]][[1]]$fit$par

results2 <- results %>%
  mutate(p.value = if_else(p.value < .001, "<0.001", as.character(round(p.value, 3))))

results2
```

Time controls for the trend, is not important but shows that rates were trending down to a stat sig degree

Time after is significant, there was a stat sig trend up during the program

Program here is not significant, so there was no evidence the program had an impact in vaccination rates compared to the US.

## Confounding variables

Comparing MN to the US would be problematic if the US had certain regions with different trends that made a national trend that was not very comparable. For example, the East Coast has lower trends because more people are vaccinated, and there is increasing vaccines in the South, since less people are vaccination, and the result of these two trends is a generally stable trend. In other words, what is a good comparison to Minnesota?

Consider comparing Minnesota with other states with similar vaccination rates (percent of population vaccinated to date).

### Vaccination rates by region

Compare regions to see if there are different rates. Haven't done this yet.

```{r}

```

## Counterfactual plot

How to plot the counterfactual? i.e. where MN would have expected vaccine rates if following the trend of the US? This may not be very empirical, but it should provide an engaging plot to see how rates differed over time.

Let's create a tidy tsibble, model both MN and USA, and compare the predictions versus the actual while the incentive was in place.

```{r}
its_tidy_ts <- its %>%
  pivot_longer(Minnesota:`United States`, names_to = "region", values_to = "vacc_rate") %>%
  select(date, prog, region, vacc_rate) %>%
  as_tsibble(key = region, index = date)

train_model = its_tidy_ts %>%
  filter(prog == 0) %>%
  model(ARIMA(vacc_rate))

train_model

its_fc <- train_model %>%
  forecast(h = 17)

its_fc %>%
  autoplot(
    its_tidy_ts
  ) +
    theme_fivethirtyeight()
```

```{r}
# extract points, turn fable -> tibble

mn_fc <- its_fc %>% 
  filter(region == "Minnesota") %>% 
  as_tibble() %>%
  full_join(its %>% select(date, Minnesota, `United States`, prog), by = "date") %>%
  select(date, `Forecasted Minnesota` = .mean, `Actual Minnesota` = Minnesota, `United States`, prog) %>%
  pivot_longer(`Forecasted Minnesota`:`United States`)

mn_fc %>%
  ggplot(aes(x = date, y = value, group = name, color = name)) +
    geom_point(shape = 1) +
    geom_smooth(data = mn_fc %>% filter(prog == 1), method = "lm", se = FALSE) +
    geom_smooth(data = mn_fc %>% filter(prog == 0), method = "lm", se = FALSE) +
    geom_vline(xintercept = as.Date("2021-07-30")) +
    theme_fivethirtyeight() +
    labs(title = "$100 Incentive and Impact on Vaccine Rates",
         subtitle = "Forecasted rates based on ARIMA model", color = "",
         caption = "@rbamattre | Source: Johns Hopkins")
```